Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-22 doc/lean.tex
CRAT --> CPOG

We set out to formally verify the system with two goals in mind: first, to ensure that the CRAT framework is mathematically
sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}). These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}. Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic. In this section, we briefly describe the functionality we implemented and what we proved about it. More information is provided in \cref{appendix:lean}.

\vspace{1em}\noindent
\textbf{Proof checking.} The goal of a CRAT proof is to construct a POG that is equivalent to the input CNF $\inputformula$. The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one. It then processes every step of the CRAT proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect. Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}. Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root.

@@

invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
\end{lstlisting}
These are maintained by valid CRAT proofs. Together with a few additional invariants
that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
% JA: I cut a few words here to save a line.

\vspace{1em}\noindent
\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
(\cref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
\begin{lstlisting}
def weightSum {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
The rules for efficient ring evaluation of partitioned formulas are expressed as:
\begin{lstlisting}
def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
\Cref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
\end{lstlisting}
To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
\lstinline{Pog.ringEval} and then proved that it matches the specification above:
\begin{lstlisting}
theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
\end{lstlisting}
Applying this to the output of our verified CRAT proof checker, which we know to be partitioned
and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
the correct ring evaluation of $\inputformula$.

>>>>

We set out to formally verify the system with two goals in mind: first, to ensure that the CPOG framework is mathematically
sound; and second, to implement correct-by-construction proof checking and ring evaluation (the ``Trusted Code'' components of \cref{fig:chain}). These two goals are achieved with a single proof development in the \lean{} programming language~\cite{demoura:cade:2021}. Verification was greatly aided by the Aesop~\cite{23limperg_aesop_white_box_best_first_proof_search_lean} automated tactic. In this section, we briefly describe the functionality we implemented and what we proved about it. More information is provided in \cref{appendix:lean}.

\vspace{1em}\noindent
\textbf{Proof checking.} The goal of a CPOG proof is to construct a POG that is equivalent to the input CNF $\inputformula$. The checker begins by parsing the input formula, initializing the set of active clauses to $\theta \leftarrow \inputformula$, and initializing the POG $P$ to an empty one. It then processes every step of the CPOG proof, either modifying its state by adding/deleting clauses in $\theta$ and adding nodes to $P$, or throwing an exception if a step is incorrect. Afterwards, it carries out the \textsc{final conditions} check of \cref{subsection:semantics}. Throughout the process, we maintain invariants which ensure that $P$ is partitioned and that a successful final check entails the logical equivalence of $\inputformula$ and $\phi_\noder$, where $\noder$ is the final POG root.

@@

invariants.partitioned : ∀ u : ILit, (st.pog.toPropForm u).partitioned
invariants.equivalent_lits : ∀ u : ILit, equivalentOver st.inputCnf.vars
    (u ∧ st.pogDefsForm) (st.pog.toPropForm x)
\end{lstlisting}
These are maintained by valid CPOG proofs. Together with a few additional invariants
that ensure the correctness of cached computations, they imply the soundness theorem for $P$ with root
node $\noder$: the \textsc{final conditions} ensure that $\theta$ is logically equivalent to
$\pogformula \doteq \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$
(\lstinline{equivalent st.clauseDb (r ∧ st.pogDefsForm)}), and from the invariants we can prove
$\inputformula \iff \phi_\noder$ (\lstinline{equivalent st.inputCnf (st.pog.toPropForm r)}).
After certifying a proof, the checker can pass its in-memory POG representation to the ring evaluator.
% JA: I cut a few words here to save a line.

\vspace{1em}\noindent
\textbf{Ring evaluation}. We formalized the central quantity (\ref{eqn:rep}) in the ring evaluation problem
(\cref{def:ring_evaluation}) in a commutative ring \lstinline{R} as follows:
\begin{lstlisting}
def weightSum {R : Type} [CommRing R]
    (weight : Var → R) (φ : PropForm Var) (s : Finset Var) : R :=
  ∑ τ in models φ s, ∏ x in s, if τ x then weight x else 1 - weight x
\end{lstlisting}
The rules for efficient ring evaluation of partitioned formulas are expressed as:
\begin{lstlisting}
def ringEval (weight : Var → R) : PropForm Var → R
  | tr       => 1
  | fls      => 0
  | var x    => weight x
  | neg φ    => 1 - ringEval weight φ
  | disj φ ψ => ringEval weight φ + ringEval weight ψ
  | conj φ ψ => ringEval weight φ * ringEval weight ψ
\end{lstlisting}
\Cref{prop:ring:eval} is then formalized as follows:
\begin{lstlisting}
theorem ringEval_eq_weightSum (weight : Var → R) {φ : PropForm Var} :
    partitioned φ → ringEval weight φ = weightSum weight φ (vars φ)
\end{lstlisting}
To efficiently compute the ring evaluation of a formula represented by a POG node, we implemented
\lstinline{Pog.ringEval} and then proved that it matches the specification above:
\begin{lstlisting}
theorem ringEval_eq_ringEval (pog : Pog) (weight : Var → R) (x : Var) :
  pog.ringEval weight x = (pog.toPropForm x).ringEval weight
\end{lstlisting}
Applying this to the output of our verified CPOG proof checker, which we know to be partitioned
and equivalent to the input formula $\inputformula$, we obtain a proof that our toolchain computes
the correct ring evaluation of $\inputformula$.

### 2023-03-22 doc/leanappendix.tex
CRAT --> CPOG

When thinking about the formal verification, it is helpful to distinguish between
the data structures that play a role in the code that is
executed and the definitions that serve as a mathematical reference,
allowing us to state our specifications and prove that they are met.
Among the former is our representation of CNF formulas:
a literal is represented as a nonzero integer, a clause is an array of literals,
and a CNF formula is an array of clauses.
\begin{lstlisting}
def ILit := { i : Int // i ≠ 0 }
abbrev IClause := Array ILit
abbrev ICnf := Array IClause
\end{lstlisting}
We define the clause database {\tt ClauseDb} as a hashmap that
stores each clause together with a flag indicating whether it has
been deleted.
Each element of a POG is either a variable, a binary disjunction (sum),
or an arbitrary conjunction (product):
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of an original variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is an extension variable appearing in the CRAT file. Note that representing edges as literals
allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
A \lstinline{Pog} is then an array of \lstinline{PogElt}
that is well-founded in the sense that each element depends only on prior elements
in the array.

@@

In this subsection, we clarify what has been verified and what has to be trusted.
Recall that our first step is to parse CNF and CRAT files in order to read in the initial
formula and the proof. We do not verify this step. Instead, the verified checker exposes
flags \verb+--print-cnf+ and \verb+--print-crat+ which reprint the consumed formula or proof,
respectively. Comparing this to the actual files using {\tt diff} provides an easy way of ensuring
that what was parsed matches their contents. This involves trusting only the correctness of the
print procedure and {\tt diff}. Similarly, if one wants to establish the correctness of the POG
contained in the CRAT file, one can print out the POG that is constructed by the checker and compare.

>>>>

When thinking about the formal verification, it is helpful to distinguish between
the data structures that play a role in the code that is
executed and the definitions that serve as a mathematical reference,
allowing us to state our specifications and prove that they are met.
Among the former is our representation of CNF formulas:
a literal is represented as a nonzero integer, a clause is an array of literals,
and a CNF formula is an array of clauses.
\begin{lstlisting}
def ILit := { i : Int // i ≠ 0 }
abbrev IClause := Array ILit
abbrev ICnf := Array IClause
\end{lstlisting}
We define the clause database {\tt ClauseDb} as a hashmap that
stores each clause together with a flag indicating whether it has
been deleted.
Each element of a POG is either a variable, a binary disjunction (sum),
or an arbitrary conjunction (product):
\begin{lstlisting}
inductive PogElt where
  | var  : Var → PogElt
  | disj : Var → ILit → ILit → PogElt
  | conj : Var → Array ILit → PogElt
\end{lstlisting}
In the first case, the argument \lstinline{x} in the expression
\lstinline{var x} is the index
of an original variable; in \lstinline{disj x left right} and \lstinline{conj x args}
it is an extension variable appearing in the CPOG file. Note that representing edges as literals
allows us to negate the arguments to \lstinline{disj} and \lstinline{conj}.
A \lstinline{Pog} is then an array of \lstinline{PogElt}
that is well-founded in the sense that each element depends only on prior elements
in the array.

@@

In this subsection, we clarify what has been verified and what has to be trusted.
Recall that our first step is to parse CNF and CPOG files in order to read in the initial
formula and the proof. We do not verify this step. Instead, the verified checker exposes
flags \verb+--print-cnf+ and \verb+--print-cpog+ which reprint the consumed formula or proof,
respectively. Comparing this to the actual files using {\tt diff} provides an easy way of ensuring
that what was parsed matches their contents. This involves trusting only the correctness of the
print procedure and {\tt diff}. Similarly, if one wants to establish the correctness of the POG
contained in the CPOG file, one can print out the POG that is constructed by the checker and compare.

### 2023-03-22 doc/mc-data.tex
Final Lean data point
CRAT --> CPOG

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CRAT file for maximum of 10,000 seconds
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \item Was able to do one-sided verification for 9 of these.
     \end{itemize}
   \item For completed CRAT files, ran {\sc crat-check} CRAT checker
     \begin{itemize}
     \item Tests all conditions for correct CRAT
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
     \end{itemize}
\end{enumerate}

\section*{Time to Generate and Validate CRAT}

@@

\section*{Generated CRAT Files}

@@

Two optimizations were implemented within the CRAT framework:
\begin{description}
\item[Lemmas:] Nodes with indegree greater than one were validated with separate lemmas, avoiding the possibly exponential expansion of the POG into a tree during the proof construction.

@@

We conducted a set of experiments using the 80 benchmark instances for
which the total number of proof clauses when both optimizations are
performed was limited to $10^7$.  We then made three additional runs
of the CRAT generator and checker, with each of the two optimizations
disabled individually and in combination.  Figures
\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:unoptimized} illustrates the effect of the two
optimizations.  Each set of points shows the number of clauses when
both optimizations are performed for the $X$ value and the number with
one or both of these optimizations disabled as the $Y$ value.

@@

The time performance is less dramatic, but still significant.
Comparing the sum of CRAT generation and checking times for each of
the four combinations of options, we get the following times relative
to the one with both optimizations enabled:
\begin{description}
\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
\end{description}

Here is another way to summarize the impact on the combined time to generate and validate the CRAT file:
\begin{itemize}
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 40 had the time reduced by at least a factor of 2
  using lemmas, 7 by using literal merging, and 49 by enabling both
  optimizations.
\end{itemize}

@@

\input{figure-mice-crat}

@@

In the figure, red dots indicate cases where neither program had an
error, but it possibly timed out.  The time-out cases are shown along
the upper edge (MICE) or the righthand edge (CRAT).

@@

Overall, we can summarize the results as:
\begin{itemize}
\item 84 total data points
\item 76 were completed by both programs in under 1000 seconds
\item Of those 21 were faster with MICE, 55 with CRAT
\item 7 completed with CRAT under 1000 seconds but exceeded that threshold for MICE
\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CRAT
\end{itemize}

@@

Ran \lean{} checker on the benchmarks yielding at most $10^7$ clauses, with time limit of 1000 seconds.
\begin{itemize}
\item 80 total
\item Had 2 that failed RUP check (need to investigate)
\item One of them needs to be rerun (Very long generation time)
\item Two timed out for \lean{}
\end{itemize}

>>>>

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CPOG file for maximum of 10,000 seconds
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \item Was able to do one-sided verification for 9 of these.
     \end{itemize}
   \item For completed CPOG files, ran {\sc cpog-check} CPOG checker
     \begin{itemize}
     \item Tests all conditions for correct CPOG
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
     \end{itemize}
\end{enumerate}

\section*{Time to Generate and Validate CPOG}

@@

\section*{Generated CPOG Files}

@@

Two optimizations were implemented within the CPOG framework:
\begin{description}
\item[Lemmas:] Nodes with indegree greater than one were validated with separate lemmas, avoiding the possibly exponential expansion of the POG into a tree during the proof construction.

@@

We conducted a set of experiments using the 80 benchmark instances for
which the total number of proof clauses when both optimizations are
performed was limited to $10^7$.  We then made three additional runs
of the CPOG generator and checker, with each of the two optimizations
disabled individually and in combination.  Figures
\ref{fig:optimized:partially-optimized}--\ref{fig:optimized:unoptimized} illustrates the effect of the two
optimizations.  Each set of points shows the number of clauses when
both optimizations are performed for the $X$ value and the number with
one or both of these optimizations disabled as the $Y$ value.

@@

The time performance is less dramatic, but still significant.
Comparing the sum of CPOG generation and checking times for each of
the four combinations of options, we get the following times relative
to the one with both optimizations enabled:
\begin{description}
\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
\end{description}

Here is another way to summarize the impact on the combined time to generate and validate the CPOG file:
\begin{itemize}
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 40 had the time reduced by at least a factor of 2
  using lemmas, 7 by using literal merging, and 49 by enabling both
  optimizations.
\end{itemize}

@@

\input{figure-mice-cpog}

@@

In the figure, red dots indicate cases where neither program had an
error, but it possibly timed out.  The time-out cases are shown along
the upper edge (MICE) or the righthand edge (CPOG).

@@

Overall, we can summarize the results as:
\begin{itemize}
\item 84 total data points
\item 76 were completed by both programs in under 1000 seconds
\item Of those 21 were faster with MICE, 55 with CPOG
\item 7 completed with CPOG under 1000 seconds but exceeded that threshold for MICE
\item 1 completed with MICE under 1000 seconds but exceeded that threshold for CPOG
\end{itemize}

@@

Ran \lean{} checker on the benchmarks yielding at most $10^7$ clauses, with time limit of 1000 seconds.
\begin{itemize}
\item 80 total
\item All ran to completion
\item All but three ran within factor of $8\times$.  Outliers were small ones
\item Harmonic mean = $5.94$.
\end{itemize}
